Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2014-12-18 doc/GMP2015/s00_abstract.tex
Most of my job for citations.

\begin{abstract}
A variety of quasi-uniform (aperiodic) and possibly self-similar, layered material structures are a natural consequence of the design objectives of rigidity and flexibility, minimizing mass, optimally distributing external stresses and participating in the assembly of diverse and multifunctional, larger  structures.

Crucial for the design and analysis of the abovementioned properties is the recursive decomposition of diverse types of underlying 2D geometric constraint systems that are isostatic or underconstrained (which we call collectively as {\it qusecs}). While the optimal recursive decomposition ({\it DR-planning}) problem is NP-hard even for 2D bar-joint constraint systems, we describe an $O(n^2)$ algorithm  for qusecs. Besides elucidating the source of NP-hardness, the algorithm outputs {\it canonical} DR-plans that meet various desirable, previously studied criteria. By leveraging recent results on Cayley configuration spaces, we show how to decompose the indecomposable algebraic systems that recombine the solutions to child subsystems at a node of an optimal DR-plan: we minimally modify the system to have a small DR-plan, leading to efficient realization algorithms. We show formal connections to well-known problems such as completion of underconstrained systems. We briefly discuss the modeling and analysis of two specific 2D layered materials as applications of the above algorithms, whose software implementation is available.

>>>>

\begin{abstract}
A variety of quasi-uniform (aperiodic) and possibly self-similar,
layered material structures are a natural consequence of design
objectives such as rigidity or flexibility, minimizing mass, optimally
distributing external stresses and participating in the assembly of
diverse and multifunctional, larger  structures.

Crucial for the design and analysis of the abovementioned properties
is the recursive decomposition of diverse types of underlying 2D
geometric constraint systems that are \vemph{isostatic or
underconstrained (which we call collectively as \dfn{qusecs})}. While
the optimal recursive decomposition (\dfn{DR-planning}) problem is NP-
hard even for (possibly overconstrained) 2D bar-joint constraint
systems we describe an $O(n^2)$ algorithm  for qusecs. Besides
elucidating NP-hardness, the algorithm outputs \dfn{canonical} DR-
plans that meet various desirable, previously studied criteria. We
leverage recent results on Cayley configuration spaces to show that
the indecomposable systems -  that are solved at node of the optimal
DR-plan by recombining solutions to child systems - can be minimally
modified to become decomposable and have a small DR-plan, leading to
efficient realization algorithms. We show formal connections to well-
known problems such as completion of underconstrained systems. We
briefly discuss the modeling and analysis of two specific 2D layered
materials as applications of the above theory and algorithms. We
intend to make a software implementation and videos available upon
request and publicly available for the final version.

### 2014-12-18 doc/GMP2015/s01_introduction.tex
Most of my job for citations.

Some properties of natural and engineered materials can be analyzed by
treating them as stacked two dimensional (2D) layers. The structure
within each layer could be self-similar \cite{Intro1} spanning
multiple scales, generally aperiodic and quasi-uniform within any one
scale, and consisting of a few repeated motifs appearing in disordered
arrangements. Each layer is not necessarily planar, i.e., it consists
of multiple, inter-constraining planar (genus 0) monolayers.
Furthermore, a layer is often  either \vemph{isostatic or
under-constrained, not self-stressed} (See Section \ref{sec:prelim} for
definitions). These properties are consistent with a self-assembled 2D
structure that minimizes mass, optimally distributes external stresses
and itself participates in the assembly of diverse and
multifunctional, larger 2D structures.

@@

\begin{enumerate}
    \item Cross-sections of microtubule structures (Figure \ref{fig:material_examples:microtubule}) \cite{Necklace1} e.g., in ciliary membranes and transitions \cite{Necklace2}

@@

    \item Crosslinked cellulose or collagen microfibril monolayers e.g., in cell-walls \cite{CellWalls1} \cite{CellWalls1}, as well as crosslinked actin filaments in the cytoskeleton matrix.

    \item More recent, engineered examples, including disordered graphene layers \cite{Graphene1} \cite{Graphene2} sometimes reinforced by  microfibrils; and DNA assemblies \cite{Microfibrils1} including a recent Sierpinski carpet, bringing other self-similar structures \cite{Microfibrils2} within reach.

    \item Silica bi-layers, glass \cite{SilicaGlass1} \cite{SilicaGlass2}, and materials that behave like assemblies of 2D particles under non-overlap constraints, e.g, jammed disks on the plane \cite{JammedDisk1}
\end{enumerate}
%
In order to study structural and mechanical properties such as
rigidity, flexes and distribution of external stresses within the
material layer, it is natural to model a material layer as a solution
of a geometric constraint system of appropriate types of geometric
primitives, under metric or algebraic constraints. Such 2D
\dfn{qusecs} (quasi-uniform or self-similar constraint systems) can be
used to understand or design material layers (their solutions) with
desired properties.

\subsection{Previous Work on Relevant 2D Geometric Constraint Systems}
We now briefly survey existing techniques for studying 2D
\dfn{qusecs},  many of which are \dfn{bar-joint} systems
(Examples 1, 2 above, see Sections \ref{sec:DRP}, \ref{sec:recomb}),
\dfn{body-hyperpin} systems (Example 4,5, see Section
\ref{sec:bodypin}) or \dfn{pinned-line incidence} systems (Example
3, see Section \ref{sec:pinnedline}). The limitations of these
techniques directly motivate the contributions of this paper.

\medskip\noindent
\header{(i) Finding (vertex)-maximal generically rigid subsystems}
Fast, graph-based algorithms exist (pebble-game \cite{Jacobs:1997:PG}), for
locating all maximal, generically rigid subsystems \uncited. When the
input is rigid, these algorithms compute the identity and output the
input.

However, both for self-similar or just aperiodic 2D qusecs, it is
imperative to recursively decompose rigid systems into their rigid
subsystems, down to the level of geometric primitives, in order to
understand or design properties at all scales, such as: rigidity,
flexes, distribution of external stresses, boundary conditions for
isostaticity, as well as behavior under constraint variations.

\medskip\noindent
\header{(ii) Optimal Recursive Decomposition (DR-planning)}
Recursive decomposition of geometric constraint systems has been
formalized \uncited and well-studied \uncited as the \dfn
{Decomposition-Recombination (DR)-planning} problem. For the above-
mentioned classes of 2D qusecs, generic rigidity is a combinatorial
property and hence each level of the decomposition should, in
principle, be achievable by a graph-based algorithm as in (1), without
involving geometric information in the constraint system. Since many
such decompositions can exist for a given constraint system, criteria
defining desirable or optimal DR-plans and DR-planning algorithms were
given in \uncited. An \dfn{optimal DR-plan} is one that minimizes
the maximum number of child subsystems of any parent system, a
property that is crucial for obtaining a solution to the parent
constraint system from the solutions of the child systems.

However, for general 2D qusecs, even when restricted to bar-joint
systems, the optimal DR-planning problem was shown to be NP-hard
\uncited.

\medskip\noindent
\header{(iii) DR-plans for special classes and with other
criteria}
For a special class of 2D qusecs, namely \dfn{tree-decomposable}
systems \uncited common in computer aided mechanical design, (which
includes ruler-and-compass and Henneberg-I constructible systems), all
DR-plans turn out to be optimal. This satisfies the so-called \dfn
{Church-Rosser} criterion, leading to highly efficient DR-planning
algorithms. For general 2D qusecs, alternate criteria were suggested
such as \dfn{cluster minimality} requiring parent systems to be
composed of a minimal set of at least 2 rigid proper subsystems (i.e.,
no proper subset forms a rigid system); and \dfn{proper maximality},
requiring child subsystems to be maximal rigid proper subsystems of
the parent system.

@@

\medskip\noindent
\header{(iv) Optimal Recombination and Solution Space Navigation}
For the entire DR-plan, finding all desired solutions is barely
tractable even if recombination of solved subsystems is easy for each
indecomposable parent system in the DR-plan. This is because even for
the simplest, highly decomposable systems with small DR-plans, the
problem of finding even a single solution to the input system at the
root of the DR-plan is NP-hard \uncited and there is a combinatorial
explosion of solutions \uncited. Typically, however, the desired
solution has a given orientation type, in which case, the crux of the
difficulty is concentrated in the algebraic complexity of
(re)combining child system solutions, to give a solution to the parent
system at any given level of the DR-plan. For fairly general 3D
constraint systems, there are algorithms with formal guarantees that
uncover underlying matroids to combinatorially obtain an optimal
parameterization to minimize the algebraic complexity of the
indecomposable parent (sub)systems that occur in the DR-plan \uncited\uncited\uncited, provided the DR-plan meets some of the abovementioned
criteria.

@@

\medskip\noindent
\header{(v) Configuration Spaces of Under-constrained Systems}
For under-constrained 2D bar-joint and body-hyperpin qusecs obtained
from various subclasses of tree-decomposable systems, algorithms have
been developed \uncited\uncited to complete them into isostatic systems
\uncited\uncited and to find paths within the connected components
\uncited\uncited of standard Cartesian configuration spaces. Most of the
algorithms with formal guarantees leverage Cayley configuration space
theory \uncited\uncited to characterize subclasses of graphs and
additional constraints that control combinatorial explosion, and
provide faithful bijective representation of connected components and
paths. These algorithms have decreasing efficiency as the subclass of
systems gets bigger, with highest efficiency for underlying partial
2-tree graphs (alternately, tree-width 2, series- parallel, $K_4$
minor avoiding), moderate efficiency for 1 degree- of-freedom (dof)
graphs with low Cayley complexity (which include common linkages such
as the Strandbeest, Limacon and Cardioid), and decreased efficiency
for general 1-dof tree-decomposable graphs. While software suites
exist \uncited\uncited, no such formal algorithms and guarantees are known
for non-tree-decomposable systems.

@@

  \item We give an efficient ($O(n^2)$) algorithm to find a canonical
  (and hence optimal) DR-plan for all 3 types of 2D qusecs mentioned
  above (Sections \ref{sec:DRP}, \ref{sec:bodypin}, and
  \ref{sec:pinnedline}). The canonical DR-plan illuminates the essence
  of the NP-hardness of finding optimal DR-plans for over-constrained
  systems.

@@

  \item Software implementation and videos can be found at:
  \todo{Provide Link}.
\end{itemize}

>>>>

Some properties of natural and engineered materials can be analyzed by
treating them as two dimensional (2D) layers. As illustrated by the
examples below, the structure within each layer is often self-
similar~\cite{2012arXiv1204.6389G} spanning multiple scales; generally
aperiodic and quasi-uniform within any one scale; and consists of a
few repeated motifs appearing in disordered arrangements. Each layer
is not necessarily planar, i.e., it consists of multiple, inter-
constraining planar (genus 0) monolayers. Furthermore, a layer is
often  either \vemph{isostatic or under-constrained, not self-
stressed} (See Section \ref{sec:prelim} for definitions). These
properties are consistent with a self-assembled 2D structure that
minimizes mass, optimally distributes external stresses and itself
participates in the assembly of diverse and multifunctional, larger 2D
structures.

@@

\begin{enumerate}
    \item Cross-sections of microtubule structures~\cite{microtubule_necklace} (Figure \ref{fig:material_examples:microtubule}) e.g., in ciliary membranes and transitions~\cite{microtubule_cilia}

@@

    \item Crosslinked cellulose or collagen microfibril monolayers e.g., in cell-walls~\cite{CellWalls1}~\cite{CellWalls1}, as well as crosslinked actin filaments in the cytoskeleton matrix.

    \item More recent, engineered examples, including disordered
    graphene layers~\cite{Graphene1}~\cite{Graphene2} sometimes
    reinforced by microfibrils; and DNA assemblies including a recent
    Sierpinski gasket~\cite{self_assembly_sierpinski}, bringing other
    self-similar structures~\cite{Microfibrils2} within reach.

    \item Silica bi-layers~\cite{silica_bilayers}, glass~\cite{sructure_of_2d_glass}, and
    materials that behave like assemblies of 2D particles under
    non-overlap constraints, i.e, like jammed disks on the plane~\cite{jammed_disks}
\end{enumerate}
%
In order to study structural and mechanical properties
of a material layer, it is natural to model a material layer as a solution or
realization of a geometric constraint system of appropriate types of geometric
primitives, under metric or algebraic constraints. Such 2D
\dfn{qusecs} (quasi-uniform or self-similar constraint systems, defined
formally in Section \ref{sec:prelim}) can be
used to understand or design material layers (their solutions) with
desired properties.

\subsection{Previous Work on Relevant 2D Geometric Constraint Systems}
We now briefly survey existing techniques for studying 2D
qusecs,  many of which are \dfn{bar-joint} systems
(Examples 1, 2 above, see Sections \ref{sec:prelim}, \ref{sec:DRP}, \ref{sec:recomb}),
\dfn{body-hyperpin} systems (Example 4,5, see Section
\ref{sec:bodypin}) or \dfn{pinned-line incidence} systems (Example
3, see Section \ref{sec:pinnedline}). The limitations of these
techniques directly motivate the contributions of this paper.

\medskip\noindent
\header{(i) Finding (vertex)-maximal generically rigid subsystems}
Fast, graph-based algorithms exist (pebble-
game~\cite{Jacobs:1997:PG}), for locating all maximal,
\dfn{generically rigid} subsystems (defined formally in Section
\ref{sec:prelim}) \uncited. When the input itself is rigid, these
algorithms do nothing, i.e., compute the identity function.

However, both for self-similar or just aperiodic 2D qusecs, it is
imperative to recursively decompose rigid systems into their rigid
subsystems, down to the level of geometric primitives, in order to
understand or design properties at all scales, such as (formally
defined in Section \ref{sec:prelim}) \dfn{rigidity}, \dfn{flexes},
distribution of \dfn{external stresses}, boundary conditions for
\dfn{isostaticity}, as well as behavior under constraint variations.

\medskip\noindent
\header{(ii) Optimal Recursive Decomposition (DR-planning).}
Recursive decomposition of geometric constraint systems has been
formalized \uncited and well-studied \uncited as the \dfn
{Decomposition-Recombination (DR)-planning} problem, defined formally in
Section \ref{sec:prelim}. For the above-
mentioned classes of 2D qusecs, generic rigidity is a combinatorial
property and hence each level of the decomposition should, in
principle, be achievable by a graph-based algorithm as in (1), without
involving geometric information in the constraint system. Since many
such decompositions can exist for a given constraint system, criteria
defining desirable or optimal DR-plans and DR-planning algorithms were
given in \uncited. An \dfn{optimal DR-plan} is one that minimizes
the \dfn{size} (defined formally in Section \ref{sec:prelim}, i.e.,
the maximum number of child subsystems of any parent system.
Being exponential the size, the complexity of solving the parent
constraint system from the solutions of the child systems, is overwhelmingly
dominated by it.

However, for general 2D qusecs that could be overconstrained, even
when restricted to bar-joint systems, the optimal DR-planning problem
was shown to be NP-hard \uncited.

\medskip\noindent
\header{(iii) DR-plans for special classes and with other
criteria.}
For a special class of 2D qusecs, namely \dfn{tree-decomposable}
systems \uncited common in computer aided mechanical design, (which
includes ruler-and-compass and Henneberg-I constructible systems), all
DR-plans turn out to be optimal. This satisfies the so-called \dfn
{Church-Rosser} criterion, leading to highly efficient DR-planning
algorithms. For general 2D qusecs, alternate criteria were suggested
such as \dfn{cluster minimality} requiring parent systems to be
composed of a minimal set of at least 2 rigid proper subsystems (i.e.,
no proper subset forms a rigid system); and \dfn{proper maximality},
requiring child subsystems to be maximal rigid proper subsystems of
the parent system. See Section \ref{sec:prelim} for formal definitions.

@@

\medskip\noindent
\header{(iv) Optimal Recombination and Solution Space Navigation.}
For the entire DR-plan, finding all desired solutions is barely
tractable even if recombination of solved subsystems is easy for each
indecomposable parent system in the DR-plan. This is because even for
the simplest, highly decomposable systems with small DR-plans, the
problem of finding even a single solution to the input system at the
root of the DR-plan is NP-hard \uncited and there is a combinatorial
explosion of solutions \uncited. Typically, however, the desired
solution has a given orientation type, in which case, the crux of the
difficulty is concentrated in the algebraic complexity of
(re)combining child system solutions, to give a solution to the parent
system at any given level of the DR-plan. For fairly general 3D
constraint systems, there are algorithms with formal guarantees that
uncover underlying matroids to combinatorially obtain an optimal
parameterization to minimize the algebraic complexity of the
indecomposable parent (sub)systems that occur in the DR-plan
\uncited\uncited\uncited, provided the DR-plan meets some of the
abovementioned criteria.

@@

\medskip\noindent
\header{(v) Configuration Spaces of Under-constrained Systems.}
For under-constrained 2D bar-joint and body-hyperpin qusecs obtained
from various subclasses of tree-decomposable systems, algorithms have
been developed \uncited\uncited to complete them into isostatic systems
\uncited\uncited and to find paths within the connected components
\uncited\uncited of standard Cartesian configuration spaces. Most of the
algorithms with formal guarantees leverage Cayley configuration space
theory \uncited\uncited to characterize subclasses of graphs and
additional constraints that control combinatorial explosion, and
provide faithful bijective representation of connected components and
paths. These algorithms have decreasing efficiency as the subclass of
systems gets bigger, with highest efficiency for underlying partial
2-tree graphs (alternately, tree-width 2, series- parallel, $K_4$
minor avoiding), moderate efficiency for 1 degree- of-freedom (dof)
graphs with low Cayley complexity (which include common linkages such
as the Strandbeest, Limacon and Cardioid), and decreased efficiency
for general 1-dof tree-decomposable graphs. While software suites
exist \uncited\uncited, no such formal algorithms and guarantees are known
for non-tree-decomposable systems.

@@

  \item We give an efficient ($O(n^2)$) algorithm to find a canonical
  (and hence optimal) DR-plan for all 3 types of 2D qusecs mentioned
  above (Sections \ref{sec:DRP}, \ref{sec:bodypin}, and
  \ref{sec:pinnedline}). The canonical DR-plan elucidates the essence
  of the NP-hardness of finding optimal DR-plans for over-constrained
  systems.

@@

  \item We intend to make software implementation and videos available
  upon request and publicly available for the final version of the
  paper.
  \todo{Provide Link}.
\end{itemize}

### 2014-12-18 doc/GMP2015/s03_drp.tex
Most of my job for citations.

The first step of the algorithm to find an optimal DR-plan is the most difficult --- to find the well-constrained vertex-maximal proper subgraphs of the input well-constrained graph. There are two approaches: (1) the bottom-up approach, using the Frontier algorithm \uncited to grow these components from single vertices; and (2) the top-down approach, using the pebble game algorithm \uncited on each node and taking the largest components that are not subgraphs of any larger ones. After the subgraphs have been found, it is a straightforward application of the theory. If the intersection of any two of the subgraphs is well-constrained, then take those two to be children of the graph. Otherwise, all the subgraphs are the children. Apply recursively to the children. This naturally terminates when you reach trivial subgraphs. The complexity of this depends on the algorithm for finding children (both are polynomial) and the depth of the plan (the deeper the plan, i.e.\ the smaller the max fan-in, the longer the running time). \todo{Think more about complexity.}

>>>>

The first step of the algorithm to find an optimal DR-plan is the most difficult --- to find the well-constrained vertex-maximal proper subgraphs of the input well-constrained graph. There are two approaches: (1) the bottom-up approach, using the Frontier algorithm \cite{Oung:2001:FFE:376957.376995} to grow these components from single vertices; and (2) the top-down approach, using the pebble game algorithm \cite{Jacobs:1997:PG} on each node and taking the largest components that are not subgraphs of any larger ones. After the subgraphs have been found, it is a straightforward application of the theory. If the intersection of any two of the subgraphs is well-constrained, then take those two to be children of the graph. Otherwise, all the subgraphs are the children. Apply recursively to the children. This naturally terminates when you reach trivial subgraphs. The complexity of this depends on the algorithm for finding children (both are polynomial) and the depth of the plan (the deeper the plan, i.e.\ the smaller the max fan-in, the longer the running time). \todo{Think more about complexity.}
